---
title: "Machine Learning vs Lightweight Deep Learning"
date: "2026-05-18"
output:
  html_document:
    toc: true
    toc_depth: 2
---

## Overview

This report summarizes the comparison between classical machine learning baselines and the lightweight deep learning pipeline. It also embeds the key visual outputs for advisor review.

```{r setup, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(fig.align = "center", out.width = "80%", dev = "svg")

comparison <- read.csv("../EKP_Database/deep_learning/experiments/experiment_comparison.csv")
ml_results <- read.csv("../EKP_Database/results/models/baseline_results.csv")
feature_importance <- read.csv("../EKP_Database/results/models/feature_importance.csv")
cor_mat <- read.csv("../EKP_Database/results/eda/correlation_analysis.csv", row.names = 1, check.names = FALSE)
features <- read.csv("../EKP_Database/interim/statistical_features/dlkcat_statistical_features.csv")
```

## Metrics Formulas (LaTeX)

$$
\mathrm{RMSE} = \sqrt{\frac{1}{n} \sum_{i=1}^{n} (y_i - \hat{y}_i)^2}
$$

$$
\mathrm{MAE} = \frac{1}{n} \sum_{i=1}^{n} \lvert y_i - \hat{y}_i \rvert
$$

$$
R^2 = 1 - \frac{\sum_{i=1}^{n} (y_i - \hat{y}_i)^2}{\sum_{i=1}^{n} (y_i - \bar{y})^2}
$$

## Overall Comparison (Test Metrics)

```{r comparison-table, echo=FALSE}
comparison
```

## Machine Learning Results

### Baseline Metrics (Valid/Test)

```{r ml-metrics, echo=FALSE}
ml_results
```

### Feature Importance

```{r rf-importance, echo=FALSE}
rf <- subset(feature_importance, model == "RandomForest")
barplot(
  rf$importance,
  names.arg = rf$feature,
  las = 2,
  cex.names = 0.8,
  main = "RandomForest Feature Importance",
  ylab = "Importance"
)
```

```{r xgb-importance, echo=FALSE}
xgb <- subset(feature_importance, model == "XGBoost")
barplot(
  xgb$importance,
  names.arg = xgb$feature,
  las = 2,
  cex.names = 0.8,
  main = "XGBoost Feature Importance",
  ylab = "Importance"
)
```

## Lightweight Deep Learning Results

### Default Run (Embedding + Mean Pooling)

```{r dl-summary-table, echo=FALSE}
comparison
```

### Ablation: CNN Encoder

```{r dl-encoder-compare, echo=FALSE}
dl_rows <- comparison[comparison$Encoder != "N/A", ]
barplot(
  dl_rows$RMSE,
  names.arg = dl_rows$Model,
  las = 2,
  cex.names = 0.8,
  main = "Deep Learning RMSE (Test)",
  ylab = "RMSE"
)
```

### Ablation: Max Pooling

```{r dl-pooling-compare, echo=FALSE}
pooling_rmse <- dl_rows[, c("Model", "Pooling", "RMSE")]
barplot(
  pooling_rmse$RMSE,
  names.arg = pooling_rmse$Model,
  las = 2,
  cex.names = 0.8,
  main = "Pooling Choice vs RMSE",
  ylab = "RMSE"
)
```

## EDA Visuals

```{r eda-corr, echo=FALSE}
cor_matrix <- as.matrix(cor_mat)
heatmap(
  cor_matrix,
  Rowv = NA,
  Colv = NA,
  scale = "none",
  main = "Correlation Heatmap"
)
```

```{r eda-target, echo=FALSE}
hist(
  features$log10_kcat,
  breaks = 40,
  main = "Target Distribution (log10 kcat)",
  xlab = "log10(kcat)"
)
```

## Feature Distribution Boxplots

```{r boxplot-protein, echo=FALSE}
protein_cols <- c(
  "protein_len",
  "protein_min",
  "protein_median",
  "protein_mean",
  "protein_std",
  "protein_max"
)
boxplot(
  features[, protein_cols],
  las = 2,
  cex.axis = 0.8,
  main = "Protein Feature Distributions"
)
```

```{r boxplot-compound, echo=FALSE}
compound_cols <- c(
  "compound_len",
  "compound_min",
  "compound_median",
  "compound_mean",
  "compound_std",
  "compound_max"
)
boxplot(
  features[, compound_cols],
  las = 2,
  cex.axis = 0.8,
  main = "Compound Feature Distributions"
)
```
